\section{Training a Small Agent}
\label{sec:training}

The \S\ref{sec:scaffold} analysis fixes the target of weight-level training: the append-only CodeAct scaffold, whose growing-prefix trajectory is the structure gradient-based objectives are built for. We report a preliminary \emph{supervised} fine-tuning study of whether training lifts a $\leq$8B agent inside that scaffold; reinforcement learning and on-policy distillation, the natural next stages, are left to future work. At matched sample size the supervised stage ties the untrained base rather than beating it --- an honest first step that locates the accuracy lever in the on-policy methods left to future work.

\subsection{What is trained, and how it is measured}

We update only the weights of the reasoning policy --- the language model that, each turn, emits a thought and a single Python action --- fine-tuning it with LoRA adapters \citep{lora} on all linear projections, which keeps a run within a single 80~GB GPU. The perceptual backbone stays frozen: the vision-language model is reached only through an external HTTP endpoint that the policy calls via \texttt{batch\_look}, and no gradient flows into it. Restricting training to the reasoning policy isolates the question of interest --- whether weight-level training can lift the agent's accuracy --- from any change in perceptual capability. The same model serves as both the policy under training and the policy under evaluation: trajectories are collected with the CodeAct scaffold, training updates the policy, and evaluation runs the identical scaffold, so the policy that is trained is exactly the policy that is deployed.

The backbone is Qwen3.5-4B, a pivot from the Qwen3-8B model named in the original proposal. The 4B model is the one that holds the $\leq$8B leaderboard slot inside this agent scaffold, making it the relevant subject for a study aimed at the $\leq$8B tier; the larger, older-generation 8B model is not the incumbent, and using it would conflate a backbone change with the training intervention. Training uses verl \citep{verl} with fully-sharded data-parallel (FSDP) model sharding.

Evaluation follows the \S\ref{sec:setup} protocol --- binary ANLS at threshold 0.9, native thinking disabled (the reasoner still writes its reasoning as free text around the code blocks, as explained in \S\ref{sec:setup}) --- applied to the 4B agent with multiple independent samples per question to reduce per-rollout variance. Two scales are used: a small 13-question stratified mini-screen (covering all eight categories) at single-sample resolution for rapid checkpoint comparison during the hyperparameter sweep, and the full 80-question validation set, evaluated at the base's own $n{=}8$ sample size, to confirm the best configuration at matched power. A model and its baseline are compared per question on the same items, so improvements are assessed with paired statistics. Three reference points anchor the trained-model numbers: the untrained Qwen3.5-4B agent in the same CodeAct scaffold, the published $\leq$8B leaderboard entry at roughly 19 ANLS, and the 8B--35B-tier entry at roughly 37.5 (a different system from the like-valued Gemini~3~Pro frontier baseline).

\subsection{Training signal}

\textbf{Rejection-sampling fine-tuning (SeqKD).} No agent reference trajectories exist, so supervised targets are generated rather than annotated. The CodeAct scaffold with a 27B teacher rolls out many trajectories over the training pool (\S\ref{sec:pool}), and only those whose submitted answer is correct under the verifier are kept --- rejection-sampling fine-tuning \citep{rft} --- after which the 4B policy is fit to the accepted trajectories with token-level cross-entropy. Because the teacher (27B) is larger than the student (4B), this is also a form of sequence-level knowledge distillation \citep{seqkd}: hard-target imitation of filtered teacher trajectories, without an auxiliary distillation loss. (Unlike self-improvement variants of rejection sampling, where a model bootstraps on its own filtered generations \citep{star,rest}, the samples here come from a separate, larger teacher rather than the policy itself.) This off-policy signal is the cheapest available and serves as a cold start: it places a frozen model onto the scaffold's turn structure and reduces the risk of a zero-reward-variance start in which no sampled rollout ever succeeds.

\textbf{Beyond supervised imitation (future work).} Supervised imitation can sharpen which answer the agent commits to but cannot raise its ceiling: at the 4B scale the agent already reaches a correct answer in some sample on far more questions than it lands consistently (pass@$k$ much exceeds single-sample accuracy; \S\ref{sec:prelim}). Converting that oracle headroom into reliable single-shot accuracy calls for an on-policy signal. The natural next stage is Group Relative Policy Optimization \citep{grpo}, which would sample a group of rollouts per prompt, score each with the answer-level ANLS verifier --- augmented with shaping terms that penalize malformed or non-executing actions --- and use the within-group reward spread as a baseline-free advantage. A per-token on-policy distillation signal --- a KL to the teacher over the student's own rollouts --- is a further option that combines imitation's density with reinforcement's distribution match \citep{opd}. Both are on-policy, and both are left to future work for the reasons noted above.

\subsection{The training-data pool}
\label{sec:pool}

DocVQA-2026 provides no training split, so supervised fine-tuning requires a substitute corpus. We assemble a pool of short-document question answering from eight public DocVQA-family datasets, each row carrying a gold answer and a rule-based verifier. Because every prompt is independently scoreable, the pool supports rejection sampling directly: of the many teacher trajectories rolled out over it, we keep only those whose submitted answer scores correct, with no human annotation of agent trajectories required.

The eight sources span the document types the benchmark evaluates: scanned business documents \citep{docvqa}, high-resolution infographics \citep{infographicvqa}, charts \citep{chartqa}, maps \citep{mapqa}, multi-page documents \citep{mpdocvqa}, numeric financial reports \citep{tatdqa}, slide decks \citep{slidevqa}, and long multi-page documents \citep{mmlongbench}. They cover five of DocVQA-2026's eight categories --- business report, infographics, maps, science poster, and slide; comics, engineering drawings, and scientific papers have no suitable short-document source and are left to transfer.

We treat document length as a first-class axis of the mixture. Most sources are single-image, so the multi-page navigation that DocVQA-2026 demands --- roughly 45\% of its validation documents span multiple pages and a third exceed ten --- would be almost absent from a naive pool. MP-DocVQA (filtered to at most three pages) and MMLongBench-Doc (median 28 pages, up to 80) supply that signal deliberately, so the pool exercises locating the right page, not only reading a single one.

The raw corpus is dominated by one source --- MapQA alone contributes 483k of its 535k questions --- so a per-source sampler draws a fixed quota from each dataset into a balanced pool of 11{,}464 prompts rather than pooling raw counts. The pool is screened for contamination against the DocVQA-2026 validation pages by perceptual-hash shortlisting followed by exact pixel comparison; the only reused images carry different questions in the two sets, so no validation answer is recoverable from the training data.

\subsection{Preliminary results}
\label{sec:prelim}

We fine-tune the 4B agent on the leakage-free teacher trajectories and sweep LoRA rank and learning rate. Training is well behaved: every configuration converges smoothly (Figure~\ref{fig:loss}), with final loss governed mainly by learning rate and only weakly by adapter rank --- the expected behaviour for LoRA fine-tuning. Because a low training loss need not imply a better agent under the observation shift of a fresh multi-turn rollout, configurations are ranked by validation accuracy rather than by loss.

On the full validation set, evaluated at the base's own $n{=}8$ sample size for a matched comparison, the best configuration --- LoRA rank 64, eight epochs --- scores 23.0 ANLS against the untrained base's 22.34 (Table~\ref{tab:sft}). The $+0.7$-point difference is small against the run-to-run spread of either model across the eight single-sample passes ($\pm 3.44$ for the base, $\pm 4.00$ for the fine-tuned agent), so the two are tied, and the oracle and consistency measures agree: pass@8 is 53.8\% against the base's 55.0\% (tied), while self-consistency over the eight samples is in fact lower (17.5\% versus 26.25\%). An intermediate single-sample reading had suggested a larger lift (28.7 ANLS), but it rested on one favorable draw, and the eight-sample average regresses to the base. The honest verdict is a null: rejection-sampling fine-tuning on teacher trajectories does not move accuracy beyond the untrained base at the 4B scale.

This is consistent with the rest of the study. Supervised imitation of teacher trajectories is off-policy, and a fresh multi-turn rollout drifts from the demonstrated observations, so imitating trajectories does not transfer to the agent's own deployment distribution --- even an in-domain set memorized to near-zero training loss only ties the untrained base. The remaining headroom is not extraction quality but completion: roughly 30\% of rollouts exhaust their turn budget before submitting on hard multi-page documents, an efficiency failure that an on-policy reward, rather than off-policy imitation, is positioned to address. Supervised fine-tuning is therefore best read as a safe warm-start that places the model on the scaffold's turn structure, with the accuracy lever lying in the on-policy methods left to future work.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lccc}
\toprule
Model & ANLS & pass@8 & SC@8 \\
\midrule
Qwen3.5-4B base & 22.34 $\pm$ 3.44 & 55.0 & 26.25 \\
\quad + SFT (rank 64, 8 ep.) & 23.0 $\pm$ 4.00 & 53.8 & 17.5 \\
\bottomrule
\end{tabular}
\caption{Best supervised configuration on the full DocVQA-2026 validation set (80Q), at matched $n{=}8$. The $\pm$ figures are the population standard deviation across the eight single-sample passes; the $+0.7$-point ANLS difference is well within either spread (a tie), pass@8 is tied, and self-consistency (SC@8) is lower. Supervised fine-tuning does not beat the untrained base at matched power.}
\label{tab:sft}
\end{table}

\input{tex/fig-loss}
